---
title: "willingness to invest in sanitation facilities"
author: "Aaron Kipngeno"
date: today
format:
  html:
    toc: true
execute:
  warning: false
editor: visual
---

```{r echo=FALSE}
#| label: load-packages
library(readr)
library(tidyverse)
```

# Project description

Data analysis to determine factors that influence households willingeness to invest in sanitation facilities in Bomet Sub-county.

objectives of the study

#1. To assess the proportion of willingness to invest in sanitation facilities by households in Bomet Sub-county.

#2. To determine the demographic and socio-cultural factors that influences the willingness to invest in sanitation facilities by households in Bomet Sub-county.

#3. To determine the economic and financial factors that influences the willingness to invest in sanitation facilities by householdsin Bomet Sub-county.

# Data Import

```{r}
sanitation <- read_csv(here::here("data/raw/Sanitation_investiment.csv")) |>
  mutate(id = row_number()) |>
  relocate(id) |> 
  janitor::clean_names()

  
```

## data cleaning

```{r}
#renaming of data frame 
sanitation_clean <- sanitation |>
mutate(id = row_number()) |>
  relocate(id) |> 
  janitor::clean_names()


sanitation_clean <- sanitation_clean |>
       rename(community_health_officers =  "source_of_information_community_health_volunteers",
      media ="source_of_information_media",
    ngo_workers = "source_of_information_ngo_workers",
  local_markertes = "source_of_information_local_marketers",
  others =  "source_of_information_others",
  quality_of_toilet = "quality_of_toilet_25",
 latrine_charcteristics = "quality_of_toilet_30",
motivating_factors = "if_yes_motivating_factors",)

sanitation_clean <- sanitation_clean |> 
  rename(public_health_officers = "source_of_information_public_health_officers"
)


#removing column
sanitation_clean$source_of_information = NULL
sanitation_clean$motivating_factors = NULL
sanitation_clean$culture= NULL
sanitation_clean$impacts_of_open_defecation = NULL

```

### converting categorical variables to factors

```{r}
sanitation_clean <- sanitation_clean |>
  mutate(age_bracket = factor(age_bracket))

sanitation_clean |>
   mutate(gender = factor(gender))

education_levels <- c("no education", "primary", "secondary", "college", "university")
sanitation_clean <- sanitation_clean |>
  mutate(education = factor(education,levels = education_levels))

religion_levels <- c("christian",  "muslim", "Other")
sanitation_clean <- sanitation_clean |>
  mutate(religion = factor(religion, levels = religion_levels))

sanitation_clean <- sanitation_clean |>
  mutate(sanitation_organizations = factor(sanitation_organizations),
         employement_status = factor(employement_status),
         willingness_to_invest = factor(willingness_to_invest),
         level_of_willingness = factor(level_of_willingness),
         knowledge_on_financial_institutions = factor(knowledge_on_financial_institutions),
         member_to_savings_and_lendings_groups = factor(member_to_savings_and_lendings_groups),
         loan = factor(loan),
         sanitation_loan = factor(sanitation_loan),
         if_no = factor(if_no),
         know_sanitation = factor(know_sanitation),
         presence_of_latrine = factor(presence_of_latrine),
         type_latrine_own = factor(type_latrine_own),
         benefit_of_use_latrine = factor(benefit_of_use_latrine),
         sanitation_organizations = factor(sanitation_organizations),
         know_sanitation = factor(know_sanitation),
         income = factor(income),
         cost_of_latrine = (cost_of_latrine),
         sanitation_laws = factor(sanitation_laws))
attach(sanitation_clean)

#checking var(class and level)
class(sanitation_clean$benefit_of_use_latrine) 
levels(sanitation_clean$benefit_of_use_latrine)

```

### data cleaning for multiple choice questions

```{r}
#1 Where do you mainly get information on sanitation?

sanitation_information <- sanitation_clean |> 
  select( public_health_officers:others) |> 
  pivot_longer(cols = public_health_officers:others,
               names_to = "information",
               values_to = "answer")
sanitation_information|> 
  filter(answer == 1) |> 
  group_by(information) |> 
  count()



#2 What are the consequences of not using latrines?(mark all that apply)

motivation <- sanitation_clean |> 
  select( disease_prevention:prestige) |> 
  pivot_longer(cols = disease_prevention:prestige,
               names_to = "consequences",
               values_to = "answer")
motivation|> 
  filter(answer == 1) |> 
  group_by(consequences) |> 
  count()

#3 What characteristics would you like to see in your latrine of choice? (mark all that apply)

latrine_charcteristics <- sanitation_clean |> 
  select( clean_floor_and_slab:lockable_door) |> 
  pivot_longer(cols = clean_floor_and_slab:lockable_door,
               names_to = "quality_of_toilet",
               values_to = "answer")
latrine_charcteristics |> 
  filter(answer == 1) |> 
  group_by(quality_of_toilet) |> 
  count()

#4What cultures do exist in your community that affect owning and use of latrines? (mark all that apply)

culture <- sanitation_clean |>
  select( in_laws_do_not_share :people_living_with_chronic_illnesses_do_not_share) |> 
  pivot_longer(cols = in_laws_do_not_share :people_living_with_chronic_illnesses_do_not_share,
               names_to = "existing_cultures",
               values_to = "answer")
culture |> 
  filter(answer == 1) |> 
  group_by(existing_cultures) |> 
  count()

#5 What hinders you/people from your community from accessing financial services?(mark all that apply)

loan_factors <- sanitation_clean |>
  select(financial_knowledge:employment_status) |> 
  pivot_longer(cols = financial_knowledge:employment_status,
               names_to = "barrier_to _loan_services",
               values_to = "answer")
loan_factors |> 
  filter(answer == 1) |> 
  group_by(`barrier_to _loan_services`) |> 
  count()
```

data explore

```{r}
type_of_latrine <- sanitation_clean |>
  filter(!is.na(type_latrine_own)) |>
  group_by(type_latrine_own) |>
  count()

latrine_access_if_dodnt_own <- sanitation_clean |> 
  filter(!is.na(if_no)) |> 
  group_by(if_no) |>
  count()
  

ggplot(data = sanitation_clean, 
       aes(x = age_bracket, fill = age_bracket))+
  geom_bar(stat = "count")+
  facet_grid(~willingness_to_invest)
```

```{r}
write.csv(sanitation_clean, "/cloud/project/data/processed/sanitation_clean.csv")
```
