---
title: "Willingness to invest in sanitation facilities. A case study of Bomet County, Kenya"
format:
  html: default
date: 02-13-2024
author:
  - name: Aaron Kipngeno
    affiliations:
      - name: OPenWASH Data Academy
execute: 
  echo: true
  warning: false
toc: true
abstract: |
  Acesss to improved sanitation is fundememtal need and  human right. However, in developing countries including Kenya, is still a challenge. this study investigated factor influncing households'willingness to invest in saniation, with majority 76% willing to invest.social factors investigated had no significant associations with the willingness to invest in sanitation, except for the benefit of use and the presence of a toilet, which showed significant associations at a p-value of less than .05. Economic factors such household income, the presence of organizations that promote sanitation, and being a member of saving and lending groups was to have significance association, at p value (<.05)
keywords:
  - sdgs
  - willingness to invest
  - social fcators
  - economic factors
citation:
       url: https://github.com/ds4owd-001/project-Aaron-Kipngeno.git
editor: visual
bibliography: citations.bib
---

```{r}
#| label: load-packages
#| code-fold: true
library(readr)
library(tidyverse)
library(knitr)
library(ggplot2)
library(usethis)
library(gtsummary)
library(ggplot2)
library(gt)
library(ggthemes)
library(tidyverse)
library(kableExtra)
```

# Introduction

The United Nations Sustainable Development Goal 6.2 aims to achieve access to adequate and equitable sanitation and hygiene for all by 2030. By, 2022, the global access to safely managed sanitation services was reported to be 57% of the population, which equates to 4.6 billion people\[ @who_notitle_2022\]. Despite this progress, over 1.5 billion people still lacked basic sanitation services, and 419 million people still practiced open defecation in Sub-Sahara Africa (SSA) and South & central Asia.

Access to improved sanitation in Kenya remains a significant challenge, with only about 32% of the population having access to basic sanitation facilities. This lack of access has led to more than 6 million Kenyans practicing open defecation, contributing to the prevalence of diseases such as diarrhea, typhoid, and cholera. In economic terms, poor sanitation leads to a loss of 0.9% of Kenya's gross domestic product (GDP) annually. In Bomet County access to improve sanitation is still low at 41% [@noauthor_dhs_2022]. The effect of poor sanitation on health in Bomet is significant, with over 40% of diarrheal illnesses and intestinal worm infestations attributed to it, as indicated by the Ministry of Health records. Investing in sanitation facilities is crucial to reducing these cases. However, many people do not prioritize sanitation on their investment agenda[@koskei_effect_2013]

The traditional community-led total sanitation (CLTS) approach, as used in Bomet, proposes that once people are made aware of their sanitation problems, they can take action to improve their sanitation status using locally available materials [@njuguna_progress_2019] whether to construc these facilities depends on people's willingness, preferences, and ability to pay for the facility of their choice. A study by [@mulatya_willingness_2021] pointed out that various factors, including demographic, socio-cultural, economic, and financial factors to have influence on Households' willingness, in rural areas of Siaya, Kakamega, and Makueni counties. the findings may not apply to Bomet due to unique contextual factors.

### Objective of the study

1\. To assess the proportion of willingness to invest in sanitation facilities by households in Bomet Sub-county.

2\. To determine the demographic and social-cultural factors that influences the willingness to invest in sanitation facilities by households in Bomet Sub-county.

3\. To determine the economic and financial factors that influences the willingness to invest in sanitation facilities by household sin Bomet Sub-county.

# Methodology

Study was Conducted in Bomet County, in Kenya. The sample size was determined using @yamane1973statistics follows: n=N/(1+N(e2) Where, n= desired sample N=Total number of households 30,086 e=sampling error (taken to be ±10%) =30, 806 / (1+30, 806 (0.12).the sample was 100 households. Cluster and simple random sampling technique was used to select households.

Data was analysed using R tool software for descriptive statistics such frequency, percentage, mean and standard deviations. Findings was presented using tables, and charts. Inferential statistics was computed using chi–square test , to examine associations of social and economic factors , and willingness to invest in sanitation.

Ethical considerations was sought in this study.

## Data Import

```{r}
sanitation <- read_csv(here::here("data/raw/Sanitation_investiment.csv"))
```

### Explore data

```{r echo=FALSE}
head(sanitation)
tail(sanitation)
nrow(sanitation)
ncol(sanitation)
```

## data cleaning

```{r, echo=FALSE}
#data transformations - rename/relocate/muatate/removing 
sanitation_clean <- sanitation |>
mutate(id = row_number()) |>
  relocate(id) |> 
  janitor::clean_names()

sanitation_clean <- sanitation_clean |>
       rename(community_health_officers =  "source_of_information_community_health_volunteers",
      media ="source_of_information_media",
    ngo_workers = "source_of_information_ngo_workers",
  local_markertes = "source_of_information_local_marketers",
  others =  "source_of_information_others",
  quality_of_toilet = "quality_of_toilet_25",
 latrine_charcteristics = "quality_of_toilet_30",
motivating_factors = "if_yes_motivating_factors",)

sanitation_clean <- sanitation_clean |> 
  rename(public_health_officers = "source_of_information_public_health_officers")

#removing column
sanitation_clean$source_of_information = NULL
sanitation_clean$motivating_factors = NULL
sanitation_clean$culture= NULL
sanitation_clean$impacts_of_open_defecation = NULL

```

### converting categorical variables to factors

```{r, echo=FALSE}
sanitation_clean <- sanitation_clean |>
  mutate(age_bracket = factor(age_bracket))

sanitation_clean |>
   mutate(gender = factor(gender))

education_levels <- c("no education", "primary", "secondary", "college", "university")
sanitation_clean <- sanitation_clean |>
  mutate(education = factor(education,levels = education_levels))

income_levels <- c("0-1000", "1001-5000", "5001-10000", "10001-20000","20001-50000", "50001")
sanitation_clean <- sanitation_clean |>
  mutate(income = factor(income, levels = income_levels))

religion_levels <- c("christian",  "muslim", "Other")
sanitation_clean <- sanitation_clean |>
  mutate(religion = factor(religion, levels = religion_levels))

sanitation_clean <- sanitation_clean |>
  mutate(sanitation_organizations = factor(sanitation_organizations),
         employement_status = factor(employement_status),
         willingness_to_invest = factor(willingness_to_invest),
         level_of_willingness = factor(level_of_willingness),
         knowledge_on_financial_institutions = factor(knowledge_on_financial_institutions),
         member_to_savings_and_lendings_groups = factor(member_to_savings_and_lendings_groups),
         loan = factor(loan),
         sanitation_loan = factor(sanitation_loan),
         if_no = factor(if_no),
         know_sanitation = factor(know_sanitation),
         presence_of_latrine = factor(presence_of_latrine),
         type_latrine_own = factor(type_latrine_own),
         benefit_of_use_latrine = factor(benefit_of_use_latrine),
         sanitation_organizations = factor(sanitation_organizations),
         know_sanitation = factor(know_sanitation),
         cost_of_latrine = (cost_of_latrine),
         sanitation_laws = factor(sanitation_laws))
attach(sanitation_clean)

#checking var(class and level)
class(sanitation_clean$benefit_of_use_latrine) 
levels(sanitation_clean$benefit_of_use_latrine)

```

### data cleaning for multiple choice questions

```{r echo=FALSE}
#| label: fig-information
#| fig-cap: "a bar grap of source of informations"
#1 Where do you mainly get information on sanitation?

sanitation_information <- sanitation_clean |> 
  select( public_health_officers:others) |> 
  pivot_longer(cols = public_health_officers:others,
               names_to = "information",
               values_to = "answer")
sanitation_information|> 
  filter(answer == 1) |> 
  group_by(information) |> 
  count() |>
  mutate(percent = n / nrow(sanitation) * 100)

# Create a bar plot for the summarized categorical data
sanitation_information |>
  filter(answer == 1) |>
  group_by(information) |>
  count() |>
  mutate(percent = n / nrow(sanitation) * 100) |>
  ggplot(aes(x = information, y = percent, fill = information)) +
  geom_bar(stat = "identity", position = position_dodge()) +
   geom_errorbar(aes(ymin = percent - 1, ymax = percent + 1), width = 0.5, color = "black") + 
  labs(title = "Source of Sanitation Information", x = "Information", y = "Percentage") +
  theme_classic()
```

```{r echo=FALSE}
#2 What are the consequences of not using latrines?(mark all that apply)
motivation <- sanitation_clean |> 
  select( disease_prevention:prestige) |> 
  pivot_longer(cols = disease_prevention:prestige,
               names_to = "consequences",
               values_to = "answer")
motivation|> 
  filter(answer == 1) |> 
  group_by(consequences) |> 
  count() |>
  mutate(percent = n / nrow(sanitation) * 100)
#3 What characteristics would you like to see in your latrine of choice? (mark all that apply)
latrine_charcteristics <- sanitation_clean |> 
  select( clean_floor_and_slab:lockable_door) |> 
  pivot_longer(cols = clean_floor_and_slab:lockable_door,
               names_to = "quality_of_toilet",
               values_to = "answer")
latrine_charcteristics |> 
  filter(answer == 1) |> 
  group_by(quality_of_toilet) |> 
  count() |> 
  mutate(percent = n / nrow(sanitation) * 100)
#4 What cultures do exist in your community that affect owning and use of latrines? (mark all that apply)
culture <- sanitation_clean |>
  select( in_laws_do_not_share :people_living_with_chronic_illnesses_do_not_share) |> 
  pivot_longer(cols = in_laws_do_not_share :people_living_with_chronic_illnesses_do_not_share,
               names_to = "existing_cultures",
               values_to = "answer")
culture |> 
  filter(answer == 1) |> 
  group_by(existing_cultures) |> 
  count() |> 
  mutate(percent = n / nrow(sanitation) * 100)
#5 What hinders you/people from your community from accessing financial services?(mark all that apply)
loan_factors <- sanitation_clean |>
  select(financial_knowledge:employment_status) |> 
  pivot_longer(cols = financial_knowledge:employment_status,
               names_to = "barrier_to _loan_services",
               values_to = "answer")
loan_factors |> 
  filter(answer == 1) |> 
  group_by(`barrier_to _loan_services`) |> 
  count() |> 
  mutate(percent = n / nrow(sanitation) * 100)
```

### Data visualization

```{r, echo=FALSE}
#| label: fig-age
#| fig-cap: "Willingeness to invest in sanitation facilities by Age Bracke "

# plot1- bar graph
ggplot(data = sanitation_clean, aes(x = age_bracket, fill = age_bracket)) +
  geom_bar(stat = "count", position = position_dodge()) +
  facet_grid(~willingness_to_invest) +
  ggtitle("Willingeness to invest in sanitation facilities by Age Bracket") +
  labs(x = "Age Bracket", y = "Count")
```

```{r echo=FALSE}
#| label: fig-scatter-plot
#| fig-cap: "scatter plot for presence of latrine, household size, willingess to invest against count"

#plot2- scatter plot
sanitation_clean |>
filter(!is.na(income)) |>
ggplot(aes(household_size))+
  geom_point( stat = "count",aes(color = presence_of_latrine), alpha = 0.5)+
  facet_grid(willingness_to_invest ~ know_sanitation)
```

```{r}
#| label: fig-boxplot
#| fig-cap: "Household Size Distribution by eduction level"
#plot- box plot
# Create a box plot
ggplot(sanitation_clean, aes(x = education, y = household_size)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Household Size Distribution by eduction level", x = "education level", y = "household size") +
  theme_classic()
```

#saving data file

```{r}
write_csv(x = sanitation_clean, "/cloud/project/data/processed/sanitation_clean.csv" )
```

# Results

importing processed data.

```{r}
#importing proccesed data

sanitation_clean <- read_csv(here::here("/cloud/project/data/processed/sanitation_clean.csv"))
```

### demographic information of the respondents

The @tbl-demographic, shows the distribution of respondents by gender, age bracket, religion, and household size. The majority of the respondents were male (49%), and the majority of the respondents' age bracket was 36-45 (30%). In terms of religion, the majority of the respondents reported being Christian (69%). The household size in Bomet County ranged from 2 to 8 persons, with a mean of approximately 4.77 and a standard deviation of 1.44. as shown in @fig-household_size.

```{r}
#| label: "tbl-demographic"
#| tbl-cap: Demographic informations of respondents.
demographic <- sanitation_clean |>
 select(gender, age_bracket, religion, education) |>
   tbl_summary(
    statistic = list(all_continuous() ~ "{mean} ± {sd}"))
 kable(demographic, caption = "Demographic informations of respondents")
```

### data visualization of demographic characteristics of respondents

```{r, echo=FALSE}
#| label: fig-pie_chart
#| fig-cap: "Distribution of respondents by gender"
#pie chart for gender distributions 
#Create a data frame with the gender distribution
gender_data <- data.frame(
  gender = c("Male", "Female", "Prefer not to say"),
  count = c(49, 48,2))

ggplot(gender_data, aes(x = "", y = count, fill = gender)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Distribution of respondents by gender", fill = "gender") +
  theme_minimal()
```

```{r, echo=FALSE}
#| label: fig-education
#| fig-cap: "Distribution of respondents bylevel of education"

#bar graph for education with error bar.
ggplot(sanitation_clean, aes(education, fill = education)) +
  geom_bar(stat = "count", fill = "skyblue") +
  labs(title = "Education Level Counts", x = "Education Level", y = "count") +
  theme_minimal()
```

```{r, echo=FALSE}
#| label: fig-religion
#| fig-cap: "Distribution of respondents by religion"
#bar graph showing religion of respondents against frequency "count"
ggplot(sanitation_clean,
       aes(x = religion, fill = religion))+
  geom_bar(stat = "count")+
  labs(title = "Frequency of Respondents by religion", x = "religion", y = "Count")

# Create a histogram for the household_size variable
# Calculate the mean, standard deviation, and size of data
mean_hs <- mean(sanitation_clean$household_size)
sd_hs <- sd(sanitation_clean$household_size)
size_hs <- nrow(sanitation_clean)
```

```{r echo=FALSE}
#| label: fig-age_bracket
#| fig-cap: "Frequency of Respondents by Age Bracket"
#bar graph showing age_bracket of respondents against frequency "count"
ggplot(sanitation_clean, aes(x = age_bracket, fill = age_bracket)) +
  geom_bar(stat = "count") +
  labs(title = "Frequency of Respondents by Age Bracket", x = "Age Bracket", y = "Count")
```

```{r}
#| label: fig-household_size
#| fig-cap: "household size distributions"

#Create a histogram for the household_size variable
ggplot(sanitation_clean, aes(x = household_size)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "orange", alpha = 0.7) +
  stat_function(fun = dnorm, args = list(mean = mean_hs, sd = sd_hs), color = "red", size = 1)+
  annotate("text", x = 1, y = 20, label = paste("Mean:", round(mean_hs, 2), "\nSD:", round(sd_hs, 2), "\nSize:", size_hs), color = "black", size = 4) +
  labs(title = "Household Size Distribution", x = "Household Size", y = "Count") +
  theme_minimal()


```

### Willingness to invest in sanitation

As illustrated in @tbl-one, majority of respondents 76% were willing to invest in sanitation facilities. The study investigated the level of willingness, 29% were not willing, 20% very much willing and 48% were willing to invest, with significant associations at p-Value (0.001) as shown in @tbl-level.

```{r}
#| label: tbl-one
#| tbl-cap: "proportion willing to invest in sanitation facilities"
tx <-sanitation_clean |> 
  tbl_summary(include = c(willingness_to_invest))

kable(tx)
```

```{r}
#| label: tbl-level
#| tbl-cap: "level of willingness"
level_of_willingness <- sanitation_clean |>
  tbl_cross(row = level_of_willingness,
            col = willingness_to_invest,
            percent = "cell") |> 
  add_p()

kable(level_of_willingness)


```

### Social factors and willingness to invest in sanitation

Study investigated associations of gender,age, religion, knowledge on satiation, presence of latrine, benefit of use of toilet, sanitation laws and willingness to invest in sanitation facilities.

#### chi-Square test statistics

A Pearson's chi-squared test was conducted to examine the relationship between Social factors and willingness to invest in sanitation.

```{r}
#gender and willinness to invest
##The results indicated a non-significant association between the two variables, χ²(df = 2, N = [99]) = 3.06, p = 0.216

s1 <- table(sanitation_clean$willingness_to_invest, sanitation_clean$gender)
chisq.test(s1)

#willingness to invest and age bracket 
#The test yielded a chi-squared statistic of 1.3313 with 4 degrees of freedom, and a p-value of 0.856. This suggests that there is no evidence to reject the null hypothesis, indicating that there is no association between the variables

s2 <- table(sanitation_clean$willingness_to_invest, sanitation_clean$age_bracket)
chisq.test(s2)

# willingess to invest & religion
#The analysis revealed no significant association, χ²(df = 2, N = 99) = 9.85, p = 0.0073
s3 <- table(sanitation_clean$willingness_to_invest, sanitation_clean$religion)
chisq.test(s3)

#willingness to invest and presence of latine
#The analysis revealed a significant association, χ²(1, N = [99]) = 13.38, p = 0.0003
s4 <- table(sanitation_clean$willingness_to_invest, sanitation_clean$presence_of_latrine)
chisq.test(s4)
# willingness to invest & benefit to use llatrine
#The analysis revealed a significant association, χ²(2, N = [99) = 19.33, p = 6.34e-05
s5 <- table(sanitation_clean$willingness_to_invest, sanitation_clean$benefit_of_use_latrine)
chisq.test(s5)

#willingess to invest in sanitation& sanitation laws
s6 <-table(sanitation_clean$willingness_to_invest, sanitation_clean$sanitation_laws)
chisq.test(s6)
```

### Economic factors and willingness to invest in sanitation

the study investigated influence of employement status, household income , sanitation organizations, knowldge on finacial institutions, and being member to saving and lending groups.

chi-square test statistics.

```{r}
#influence of employement status
#no significant association p value = 0.02 > 0.005
e1 <- table(sanitation_clean$willingness_to_invest, sanitation_clean$employement_status)
chisq.test(e1)

# influnce of household income
#The analysis revealed a significant association, χ²(4, N = [99]) = 16.22, p = 0.0027
e2 <- table(sanitation_clean$willingness_to_invest, sanitation_clean$income)
chisq.test(e2)

#presence sanitation organizations
#the analysis revealed a significant association, χ²(1, N = [99]) = 12.07, p = 0.0005
e3 <- table(sanitation_clean$willingness_to_invest, sanitation_clean$sanitation_organizations)
chisq.test(e3)
# influence of being member of saving and lending groups
# signifiacnt association at χ²(1, N= 99)= 10.692, p = 001. 
e4<- table(sanitation_clean$willingness_to_invest, sanitation_clean$member_to_savings_and_lendings_groups)
chisq.test(e4)

```

# Conclusions

The study found that the majority of respondents, 76%, were willing to invest in sanitation. There could be several reasons for this willingness, including the desire for personal safety and privacy, fear of penalties and fines, fear of disease outbreaks, raising social status, and the need for a safe place for defecation/urination.

The study found that the majority of social factors investigated had no significant associations with the willingness to invest in sanitation, except for the benefit of use and the presence of a toilet, which showed significant associations at a p-value of less than .005. This could be due to various reasons, including the specific emphasis on the benefit of use and the presence of a toilet in influencing individuals' decisions to invest in sanitation. Additionally, the findings may suggest that these two factors, in particular, play a more prominent role in shaping attitudes and behaviors towards sanitation investment compared to other social factors that were examined in the study

Several economic factors were found to have significant associations with the willingness to invest in sanitation, including household income, the presence of organizations that promote sanitation, and being a member of saving and lending groups at p value (\<.005). This could be due to increasing access to financial resources and promoting economic development could help to improve sanitation investment.

# References
